---
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

set.seed(1)
```

# randomPlantedForest

<!-- badges: start -->
[![R-CMD-check](https://github.com/PlantedML/randomPlantedForest/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PlantedML/randomPlantedForest/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PlantedML/randomPlantedForest/graph/badge.svg)](https://app.codecov.io/gh/PlantedML/randomPlantedForest)
[![randomPlantedForest status badge](https://plantedml.r-universe.dev/badges/randomPlantedForest)](https://plantedml.r-universe.dev/randomPlantedForest)
<!-- badges: end -->

`randomPlantedForest` implements Random Planted Forest
([Hiabu, Mammen & Meyer][arxiv1]), a tree ensemble whose predictions you can
read off directly, without post-hoc explanation methods.

Like a random forest, it averages many tree-based models grown on bootstrap
samples. Unlike a random forest, each of these models is a sum of trees that
each split on a fixed set of predictors, and `max_interaction` bounds how many
predictors that can be. With `max_interaction = 1` the forest is an additive
model; with `2` it adds pairwise interactions, and so on. As a result, the
fitted model decomposes exactly into an intercept, main effects and
interactions up to that order:

$$\hat m(x) = \hat m_0 + \sum_k \hat m_k(x_k) + \sum_{k < l} \hat m_{kl}(x_k, x_l) + \dots$$

Each component can be inspected and plotted on its own, and together they sum
to the prediction.

## Installation

Install the development version from [r-universe][r-universe] with

```r
install.packages("randomPlantedForest", repos = "https://plantedml.r-universe.dev")
```

or from [GitHub](https://github.com/PlantedML/randomPlantedForest) with

```r
# install.packages("pak")
pak::pak("PlantedML/randomPlantedForest")
```

## Example

`rpf()` takes a formula, x/y data or a [recipe](https://recipes.tidymodels.org/),
and `predict()` returns a tibble as in tidymodels:

```{r helloworld}
library(randomPlantedForest)

mtcars$cyl <- factor(mtcars$cyl)
rpfit <- rpf(mpg ~ cyl + wt + hp, data = mtcars, ntrees = 25, max_interaction = 2)
rpfit

head(predict(rpfit, new_data = mtcars))
```

`predict_components()` returns the decomposition: one column per main effect
and interaction, plus the intercept.

```{r components}
components <- predict_components(rpfit, new_data = mtcars)
head(components$m)
```

The [glex][glex] package plots these components:

```{r effects, message = FALSE}
library(glex)
library(ggplot2)
library(patchwork)

(autoplot(components, "wt") + autoplot(components, "hp")) /
  (autoplot(components, "cyl") + autoplot(components, c("wt", "hp")))
```

The [Get started][get-started] guide covers classification, recipes, saving
models and more on interpreting components, and the
[Bikesharing decomposition][bikes] article works through a larger example.

[arxiv1]: https://arxiv.org/abs/2012.14563
[get-started]: https://plantedml.com/randomPlantedForest/articles/randomPlantedForest.html
[bikes]: https://plantedml.com/glex/articles/Bikesharing-Decomposition-rpf.html
[r-universe]: https://plantedml.r-universe.dev/packages
[glex]: https://plantedml.com/glex/
