---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

set.seed(1)
```

# randomPlantedForest

<!-- badges: start -->
[![R-CMD-check](https://github.com/PlantedML/randomPlantedForest/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PlantedML/randomPlantedForest/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PlantedML/randomPlantedForest/branch/master/graph/badge.svg)](https://app.codecov.io/gh/PlantedML/randomPlantedForest?branch=master)
<!-- badges: end -->

`randomPlantedForest` implements "Random Planted Forest", a directly interpretable tree ensemble [(arxiv)][arxiv1].

## Installation

You can install the development version of `randomPlantedForest` from [GitHub](https://github.com/) with:

```r
# install.packages("remotes")
remotes::install_github("PlantedML/randomPlantedForest")
```

## Example

Model fitting uses a familiar interface:

```{r}
library(randomPlantedForest)

mtcars$cyl <- factor(mtcars$cyl)
rpfit <- rpf(mpg ~ cyl + wt + hp, data = mtcars, ntrees = 25, max_interaction = 2)
rpfit

predict(rpfit, new_data = mtcars) |>
  cbind(mpg = mtcars$mpg) |>
  head()
```

Prediction components can be accessed via `predict_components`, including the intercept, main effects, and interactions up to a specified degree.
The returned object also contains the original data as `x`, which is required for visualization.
Alternative, the `glex` package can be used, but `glex(rpfit)` would yield the same result.

```{r}
components <- predict_components(rpfit, new_data = mtcars) 

str(components)
```

Various visualization options are available vai `glex`, e.g. for main and second-order interaction effects:

```{r}
# install glex if not available:
if (!requireNamespace("glex")) remotes::install_github("PlantedML/glex")
library(glex)
library(ggplot2)
library(patchwork) # For plot arrangement

p1 <- autoplot(components, "wt")
p2 <- autoplot(components, "hp")
p3 <- autoplot(components, "cyl")
p4 <- autoplot(components, c("wt", "hp"))

(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Selected effects for mtcars",
    caption = "(It's a tiny dataset but it has to fit in a README, okay?)"
  )
```


See the [Bikesharing decomposition][bikes] article for more examples.


[arxiv1]: https://arxiv.org/abs/2012.14563
[bikes]: https://plantedml.com/glex/articles/Bikesharing-Decomposition-rpf.html
