---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# randomPlantedForest

<!-- badges: start -->
[![R-CMD-check](https://github.com/PlantedML/randomPlantedForest/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PlantedML/randomPlantedForest/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PlantedML/randomPlantedForest/branch/master/graph/badge.svg)](https://app.codecov.io/gh/PlantedML/randomPlantedForest?branch=master)
<!-- badges: end -->

`randomPlantedForest` implements "Random Planted Forest", a directly interpretable tree ensemble [(arxiv)][arxiv1].

## Installation

You can install the development version of `randomPlantedForest` from [GitHub](https://github.com/) with:

```r
# install.packages("remotes")
remotes::install_github("PlantedML/randomPlantedForest")
```

## Example

Model fitting uses a familiar interface:

```{r}
library(randomPlantedForest)

rpfit <- rpf(mpg ~ cyl + wt + hp, data = mtcars, ntrees = 25, max_interaction = 2)
rpfit

predict(rpfit, new_data = mtcars) |>
  cbind(mpg = mtcars$mpg)
```

Prediction components can be accessed via `extract_components`, including the intercept, main effects, and interactions up to a specified degree:

```{r}
m <- extract_components(rpfit, new_data = mtcars) 
m
```



[arxiv1]: https://arxiv.org/abs/2012.14563
