---
title: "Getting Started with randomPlantedForest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with randomPlantedForest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

`randomPlantedForest` implements Random Planted Forest ([Hiabu, Mammen & Meyer](https://arxiv.org/abs/2012.14563)), a directly interpretable tree ensemble.
Its defining feature is that the maximum order of interaction between predictors can be bounded via `max_interaction`: the fitted model then decomposes exactly into main effects, two-way interactions, and so on, a functional ANOVA decomposition, with no post-hoc approximation needed.

```{r setup}
library(randomPlantedForest)
```

## Fitting a regression forest

To see what the model recovers, we simulate data where the true components are known.
The outcome depends on `x1`, `x2` and `x3` through main effects of different shapes and on `x1` and `x2` jointly through a product interaction with `x4` having no effect at all:

$$y = 1 + \sin(\pi x_1) + 2\left(x_2^2 - \tfrac{1}{3}\right) + \tfrac{x_3}{2} + 2 x_1 x_2 + \varepsilon$$

With uniform predictors on $[-1, 1]$, each term has mean zero and the interaction has no main-effect part, so the terms are exactly the components the model is meant to estimate.

```{r simulate}
set.seed(1)
n <- 1000
sim <- data.frame(
  x1 = runif(n, -1, 1),
  x2 = runif(n, -1, 1),
  x3 = runif(n, -1, 1),
  x4 = runif(n, -1, 1)
)
true_effects <- list(
  x1 = function(x) sin(pi * x),
  x2 = function(x) 2 * (x^2 - 1 / 3),
  x3 = function(x) x / 2
)
sim$y <- 1 +
  true_effects$x1(sim$x1) +
  true_effects$x2(sim$x2) +
  true_effects$x3(sim$x3) +
  2 * sim$x1 * sim$x2 +
  rnorm(n, sd = 0.3)
```

```{r fit-regression}
rpfit <- rpf(y ~ ., data = sim, max_interaction = 2, ntrees = 50)
rpfit
```

`max_interaction = 2` restricts the forest to main effects and two-way interactions.
This is the central tuning choice of the method: lower values give a more interpretable, more constrained model, while `max_interaction = 0` allows interactions of any order.
`ntrees` sets the number of tree families in the ensemble.

Prediction works as usual:

```{r predict-regression}
predict(rpfit, new_data = head(sim))
```

Note that `rpf()` does not handle missing values: rows containing `NA` must be imputed or removed beforehand (see the recipes section below for one way to do this as part of the model specification).

## Decomposing predictions

Because the interaction order is bounded, the prediction function is a sum of low-dimensional components:

$$\hat{m}(x) = \hat{m}_0 + \sum_k \hat{m}_k(x_k) + \sum_{k < l} \hat{m}_{kl}(x_k, x_l)$$

`predict_components()` returns exactly this decomposition: an intercept and one column per main effect and interaction term, evaluated on the supplied data.
The forest is *purified* internally to make the decomposition unique, which is why the print method above reports the purification state.

```{r components}
components <- predict_components(rpfit, new_data = sim)

components$intercept
names(components$m)
```

The components sum exactly to the model prediction:

```{r components-sum}
pred_sum <- components$intercept + rowSums(components$m)
pred_direct <- predict(rpfit, new_data = sim)$.pred

all.equal(pred_sum, pred_direct)
```

Each column of `components$m` is the isolated contribution of one predictor (or one pair) to the prediction, on the scale of the outcome.
Their standard deviations give a simple importance measure: the three main effects and the `x1:x2` interaction dominate, while `x4` and the other interactions are close to zero, as they should be.

```{r components-sd}
round(sort(sapply(components$m, sd), decreasing = TRUE), 3)
```

The estimated main effects follow the true functions (blue lines).
As with any tree-based model, they flatten out towards the edges of the data, where few observations are available to determine them.

```{r main-effects, fig.height = 3}
op <- par(mfrow = c(1, 3), mar = c(4, 4, 1, 1))
for (v in names(true_effects)) {
  ylim <- range(components$m[[v]], true_effects[[v]](sim[[v]]))
  plot(
    sim[[v]],
    components$m[[v]],
    pch = 20,
    col = "darkgray",
    ylim = ylim,
    xlab = v,
    ylab = "component"
  )
  curve(true_effects[[v]](x), add = TRUE, lwd = 2, col = "blue")
}
par(op)
```

The interaction is a function of two variables, so it is shown as a heatmap.
We evaluate the estimated `x1:x2` component on a grid. Note `x3` and `x4` must be present in the new data but do not affect this component.

```{r interaction, fig.height = 3.5}
grid_points <- seq(-1, 1, length.out = 50)
grid <- expand.grid(x1 = grid_points, x2 = grid_points, x3 = 0, x4 = 0)
estimated <- predict_components(rpfit, new_data = grid, predictors = c("x1", "x2"))$m[["x1:x2"]]

surfaces <- list(
  true = outer(grid_points, grid_points, function(x1, x2) 2 * x1 * x2),
  estimated = matrix(estimated, nrow = length(grid_points))
)
zlim <- c(-1, 1) * max(abs(unlist(surfaces)))

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (s in names(surfaces)) {
  image(
    grid_points,
    grid_points,
    surfaces[[s]],
    zlim = zlim,
    col = hcl.colors(51, "Blue-Red 3"),
    xlab = "x1",
    ylab = "x2",
    main = paste(s, "x1:x2")
  )
}
par(op)
```

The plots above are drawn manually from `predict_components()` for illustration.
The [`glex`](https://plantedml.com/glex/) package provides `ggplot2` versions of them: `glex()` computes the decomposition for a fitted `rpf` object, and `plot_main_effect()` and `plot_twoway_effects()` draw main effects and two-way interactions.

## Classification

For a factor outcome, `rpf()` automatically fits a classification forest.
We illustrate with the `penguins` data from `palmerpenguins`, removing missing values first.

### Binary

```{r classif-binary}
penguins <- na.omit(palmerpenguins::penguins)

rpfit_sex <- rpf(sex ~ ., data = penguins, loss = "exponential", ntrees = 50)

predict(rpfit_sex, head(penguins), type = "prob")
predict(rpfit_sex, head(penguins), type = "class")
```

With `loss = "exponential"` (or `"logit"`), `type = "link"` additionally gives raw predictions on the log-odds scale.

### Multiclass

Multiclass classification works the same way: probability predictions contain one column per class:

```{r classif-multiclass}
rpfit_species <- rpf(species ~ ., data = penguins, ntrees = 50)

predict(rpfit_species, head(penguins), type = "prob")
predict(rpfit_species, head(penguins), type = "class")
```

The default loss is `"L2"`.
 `"L1"`, `"logit"` and `"exponential"` are also available.
Note that with `loss = "logit"`, multiclass models are fit on a reference-class (log-odds) scale, so link predictions and components have one column less than there are classes: see `?predict.rpf` for details.

## Using recipes

`rpf()` also accepts a [recipes](https://recipes.tidymodels.org/) recipe in place of a formula, so preprocessing can be bundled with the model specification.
This makes it easy to add preprocessing steps, such as imputing missing values, which `rpf()` itself does not accept:

```{r recipes, message = FALSE}
library(recipes)

# rows with a missing outcome still need to be dropped by hand
peng_raw <- subset(palmerpenguins::penguins, !is.na(body_mass_g))

rec <- recipe(body_mass_g ~ ., data = peng_raw) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors())

rpfit_rec <- rpf(rec, data = peng_raw, ntrees = 50)

predict(rpfit_rec, head(peng_raw))
```

The preprocessing steps are stored with the model and applied automatically in `predict()`.

## Saving and restoring models

An `rpf` object holds its forest in an external pointer to a C++ object, which `saveRDS()` cannot preserve, meaning that a naively saved and restored model is not usable:

```{r saveRDS-trap, error = TRUE}
tmp <- tempfile(fileext = ".rds")
saveRDS(rpfit, tmp)
broken <- readRDS(tmp)

rpf_is_valid(broken)
predict(broken, head(sim))
```

Use `rpf_marshal()` to convert the model into a plain R list before saving, and `rpf_unmarshal()` after loading:

```{r marshal}
saveRDS(rpf_marshal(rpfit), tmp)
restored <- rpf_unmarshal(readRDS(tmp))

all.equal(
  predict(restored, head(sim)),
  predict(rpfit, head(sim))
)
```

Alternatively, a [bundle](https://rstudio.github.io/bundle/) method wrapping the same mechanism is available via `bundle::bundle()` and `bundle::unbundle()`.

```{r cleanup, include = FALSE}
unlink(tmp)
```
